Detection and replication of epistasis influencing transcription in humans

### 2013-06-26 docs/manuscript/methods.tex
more work

We tested for enrichment of genomic features for the 687 IndexSNPs that comprise the 434 epistatic interactions with data present in discovery and replication datasets. For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$. We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19, GENCODE annotation version 16 \cite{Harrow2012}) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions. Therefore a particular IndexSNP might cover multiple genomic features.

>>>>

We tested for enrichment of genomic features for the 687 IndexSNPs that comprise the 434 epistatic interactions with data present in discovery and replication datasets. For each of the 687 IndexSNPs we calculated LD with all regional SNPs within a radius of 0.5Mb and kept all regional SNPs with LD $r^{2} > 0.8$. We then cross-referenced the remaining regional SNPs with the annotated chromatin structure reference \cite{Hoffman2012}) querying whether the regional SNPs fell in Predicted promoter region including TSS (TSS), Predicted promoter flanking region (PF), Predicted enhancer (E), Predicted weak enhancer or open chromatin cis regulatory element (WE), CTCF enriched element (CTCF), Predicted transcribed region (T), or Predicted Repressed or Low Activity region (R) positions. Therefore a particular IndexSNP might cover multiple genomic features.

### 2013-06-26 docs/manuscript/nature_letter.tex
more work

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes. Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

@@

This work could not have been completed without access to high performance GPGPU compute clusters. We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au). We also acknowledge Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{OLapWithGenomicFeaturesFinal}
	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1kb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions. Green bars represent the proportion (\emph{y}-axis) of the 528,509 SNPs used in the analysis that fall within the range of the different genomic features. There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features.}
	\label{fig:snplocations}
\end{figure}
\clearpage

@@

\begin{ThreePartTable}
\begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}

>>>>

% \subsection{Initial search in discovery set and replication}
In our discovery dataset (Brisbane Systems Genetics Study, BSGS \cite{Powell2012}) of 846 individuals genotyped at 528,509 SNPs, we exhaustively tested every pair of SNPs for genetic interactions against each of 7339 expression traits in whole blood. After stringent filtering and multiple testing correction (Methods) we identified 501 putative genetic interactions influencing the expression levels of 238 genes (Supplementary Table \ref{tab:results}). Of the 501 discovery interactions, 434 had available data and passed filtering (Methods) in two independent replication datasets, Fehrmann \cite{Fehrmann2011} and the Estonian Genomics Centre University of Tartu (EGCUT) \cite{Metspalu2004}, in which we saw convincing evidence for replication. We used the summary statistics from the replication datasets to perform a meta analysis to obtain an independent $p$-value for the putative interactions, and 30 were significant after applying a Bonferroni correction for multiple testing (Table \ref{tab:bonferroni}). These significant interactions exhibited remarkable similarity in GP maps between all three datasets (Figure \ref{fig:gpmaps}).

@@

This work could not have been completed without access to high performance GPGPU compute clusters. We acknowledge iVEC for the use of advanced computing resources located at iVEC@UWA (www.ivec.org), and the Multi-modal Australian ScienceS Imaging and Visualisation Environment (MASSIVE) (www.massive.org.au). We also thank Jake Carroll and Irek Porebski from the Queensland Brain Institute Information Technology Group for HPC support.

@@

\begin{figure}
	\centering
	\includegraphics[width=5in]{ChromatinStatesOverlap}
	\caption{\textbf{Location of SNPs relative to genomic features} All SNPs within 1Mb and $r^2 > 0.8$ of each \emph{cis}- and \emph{trans}-SNP were taken to find which genomic features (\emph{x}-axis) were covered by the SNPs that compose the 501 significant interactions. Green bars represent the proportion (\emph{y}-axis) of the 528,509 SNPs used in the analysis that fall within the range of the different genomic features. There is enrichment for \emph{cis}-acting SNPs (red bars) in promotor regions, but \emph{trans}-acting SNPs (blue bars) are not enriched for genomic features. The labels on the $x$-axis are as follows: TSS = Predicted promoter region including TSS, PF = Predicted promoter flanking region, E = Predicted enhancer, WE = Predicted weak enhancer or open chromatin cis regulatory element, CTCF = CTCF enriched element, T = Predicted transcribed region, R = Predicted Repressed or Low Activity region}
	\label{fig:snplocations}
\end{figure}
\clearpage

@@

\begin{ThreePartTable}
\label{tab:results}
\begin{longtable}{|llr|lrrl|lrrl|rrrr|r|}
